---
title: "Getting started with imurun"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with imurun}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

imurun is a beginner-friendly, command-line front-end to the
[imuGAP](https://github.com/ACCIDDA/imuGAP) vaccine-coverage model. You describe
your data in a spreadsheet; imurun validates it, fits the model, and writes back
coverage estimates. No R knowledge is required.

This guide walks the whole happy path: install, try the shipped example,
prepare your own workbook, validate, fit, and read the results.

## Install

```sh
Rscript -e 'remotes::install_github("ACCIDDA/imurun")'
```

Once imurun is on CRAN you will be able to `install.packages("imurun")`.

To call `imurun` as a bare command on your `PATH` (instead of typing
`Rscript -e ...` every time), follow the one-time CLI setup in
[issue #16](https://github.com/ACCIDDA/imurun/issues/16). The commands below are
written as if `imurun` is on your `PATH`.

## 1. Run the shipped example first

imurun ships a complete, ready-to-run example so you can see a fit before
preparing your own data. Write a copy into the current folder:

```sh
imurun example .        # writes imurun_example.xlsx
```

The example is a real, fittable dataset carved from imuGAP's simulated data.
From R, its installed location is:

```{r}
imurun::imurun_example()
```

Open it in your spreadsheet program. The first tab is **instructions**; then
come **observations**, **locations**, and an optional **target** tab.

## 2. Prepare your own workbook

Start from a blank, correctly-headed template:

```sh
imurun init .           # writes imurun_template.xlsx
```

Fill the tabs (the instructions tab explains every column). imurun uses plain,
human-readable headers:

- **observations** -- one row per sampled count: *Location*, *Reference
  cohort*, *Youngest age*, *Oldest age*, *Dose*, *Vaccinated*, *Sampled* (and
  an optional *Censored*).
- **locations** -- the location hierarchy: *Location* and *Parent location*
  (leave *Parent location* blank for the single top-level location).
- **target** (optional) -- the populations you want coverage predicted for:
  *Location*, *Reference cohort*, *Youngest age*, *Oldest age* (and optional
  *Dose* and *Label*).

Two things imurun handles for you, so you never fill them in:

- there is **no observation id** to invent -- imurun assigns one; and
- there is **no populations tab** -- imurun builds populations from your
  observations.

If you prefer imuGAP's own column names (`loc_id`, `cohort`, ...), those work
too; and you may delete the instructions tab -- imurun only reads the data
sheets.

### Counts that cover a range of ages

A count is often drawn from a band of ages rather than a single one, so an
observation names an age range. For a single-age count put the same age in
*Youngest age* and *Oldest age*, or simply use one *Age* column instead --
imurun reads that as a one-age range.

When a count does span several ages, imurun splits it evenly across them: each
age contributes an equal share of that observation's modeled coverage. The
sheet carries no per-age denominators, so an even split is the only division
the input can express; weighting by age-specific population sizes would need a
further column and is not available yet.

*Reference cohort* is the cohort of the **oldest** age in the row. Younger ages
in the same row belong to correspondingly later cohorts, so a row describes one
moment in time rather than one cohort followed through life. The *target* sheet
reads its age span exactly the same way, which is what lets you write a target
over the same span as an observation and get back an estimate of the same
thing.

## 3. Validate before you fit

Check a workbook without fitting anything:

```sh
imurun -h imurun_example.xlsx
```

From R, that is `read_inputs()` followed by `validate_inputs()`:

```{r}
inputs <- imurun::read_inputs(imurun::imurun_example())
imurun::validate_inputs(inputs)
cat("Looks good -- ready to fit.\n")
```

When something is wrong, imurun reports it in plain, spreadsheet terms rather
than an R error. For instance, if a *Sampled* cell holds text instead of a
number:

```{r, error = TRUE}
broken <- inputs
broken$obs$sample_n[1] <- "not a number"
imurun::validate_inputs(broken)
```

Fix the flagged cell(s) in your spreadsheet and validate again.

## 4. Fit the model

```sh
imurun imurun_example.xlsx
```

This fits the imuGAP model and writes two files next to your input:

- **`results.xlsx`** -- a copy of your workbook with a `results` sheet added.
  This is the one most people want.
- **`fit.rds`** -- the raw fitted model, for post-processing in R.

Fitting compiles and runs a Stan model, so it can take several minutes. It is
not run in this guide; on success the command exits with status `0`.

imurun will not overwrite existing output: a second run into the same place
stops with an error rather than replacing results you may not have read yet.
Pass `--overwrite` when you do want to replace them. Two other flags control
where the output goes:

```sh
# put the amended workbook somewhere specific, and write a results-only CSV too
imurun imurun_example.xlsx --results estimates.xlsx --csv estimates.csv
```

Your target sheet is checked before fitting starts, so a typo in a target row
fails immediately instead of after a long fit.

## 5. Read the results

Open `results.xlsx` and go to the **results** sheet. It has one row per
requested population (each location, at each age in the span you asked for):

| Column | Meaning |
|:---|:---|
| `target_id` | your label from the target sheet, if you gave one |
| `loc_id`, `cohort`, `age`, `dose` | which population the row estimates |
| `est_median` | the estimated coverage |
| `est_lower`, `est_upper` | the credible interval around it |
| `ci_level` | the interval's level (0.95) |
| `n_draws` | posterior draws behind the estimate |

Your original `observations`, `locations`, and `target` sheets are copied over
unchanged, so the request travels with the answer and your input file is left
untouched.

`fit.rds` is the raw fitted model, for anyone who wants to post-process it in R:

```{r, eval = FALSE}
fit <- readRDS("fit.rds")
```

## Where next

- **Run `imurun` as a command** (PATH setup, or the `Rscript -e` fallback):
  [issue #16](https://github.com/ACCIDDA/imurun/issues/16).
- **The model itself**: see the [imuGAP](https://github.com/ACCIDDA/imuGAP)
  documentation.
